Missing the Point$\colon$ Non-Convergence in Iterative Imputation Procedures

### 2020-04-15 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Rewrote intro for peer feedback research seminar.

---
title: Missing the Point$\colon$ Non-Convergence in Iterative Imputation Procedures
runninghead: Oberman
author:
- name: H. I. Oberman
  num: 1
address:
- num: 1
  org: Department of Methodology and Statistics, Utrecht University, Utrecht, The Netherlands
corrauth: "Hanne Oberman, Sjoerd Groenman building, Utrecht Science Park, Utrecht, The Netherlands."
email: h.i.oberman@uu.nl
abstract: "(**Rewrite:**) Multple imputation by chained equations (mice) is a widely used tool to accommodate missing data. While empirical evidence supports the validity of inferences obtained using mice, there is no consensus on the convergence properties of the method. This paper provides insight into non-convergence of mice algorithms."
keywords: MICE; convergence 
classoption:
  - Royal
  - times
bibliography: thesis.bib
bibliographystyle: sageh
output:
  rticles::sage_article:
    keep_tex: yes
---

@@

@@

<!-- (**or** Everybody who works with person data is likely to run into a missing data problem.) -->
Missing data problems are ubiquitous and pervasive in the social and behavioral sciences. If a dataset contains just one missing value for a variable of interest, statistical inferences are undefined and will not produce any result. Incomplete observations are therefore ignored by default in many statistical packages (i.e., list-wise deletion is employed). Unfortunaltely, this *ad hoc* solution may yield wildly invalid results [@buur18]. An alternative is to 'impute' (i.e., fill in) every missing value in an incomplete dataset. With imputation techniques, one or several completed datasets are created, on which statistical inferences can be performed. The case with several completed datasets is known as 'multiple imputation' [MI; @rubin76], and requires an additional step to pool the results accross imputations. By applying Rubin's rules, the pooled result will reflect the uncertainty in the data due to missingness. Under many circumstances, this yields an unbiased and confidence valid estimate of the inference on the true--but missing--data [@buur18]. A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'. MICE is an iterative algorithmic procedure to draw imputations from the posterior predictive distribution of the missing values. This introduces a potential threat to the validity of the imputations: what if the algorithm has not converged? Are the implications then to be trusted? And can we rely on the inference obtained on the completed data? These are all open questions, because the convergence properties of iterative imputation algortihms have not been systematically studied [@taka17; @buur18]. In this paper, we investigate the convergence properties of iterative imputation procedures by means of model based simulation in `R` [@R].

@@

<!-- by which  That is, unless ad hoc solutions are employed (e.g., list-wise deletion), which can yield invalid results.  contains missing values, statistical analyses cannot be performed without  only be performed a dataset contains missing values, statistical analyses can statistical inferences cannot be performed   -->
<!-- <!-- At some point, any social scientist  --> -->
<!-- <!-- conducting statistical analyses  --> <!-- will run into a missing data problem [@alli02].  --> -->
<!-- Missingness is problematic because statistical inference cannot be performed on the whole data without employing \emph{ad hoc} solutions (e.g., list-wise deletion), which may yield wildly invalid results [@buur18]. A popular method to obtain statistically valid results despite of missing data is 'multiple imputation' [MI; @rubin76]. The multiple imputation method can be roughly divided into 3 steps: First, each missing value in an incomplete dataset is 'imputed' (i.e., filled in) several times by some algorithmic procedure. Next, the analysis of scientific interest is performed on each of the completed datasets. And finally, the results of these analyses are pooled into a single aggregated statistic.  -->

@@

<!-- The theoretical foundation of these streps was laid by Rubin [-@rubin76, -@rubin87]. For an accessible and comprehensive introduction to MI from an applied perspective, see e.g. @buur18. In this paper, we focus on the first of these steps: the algorithmic procedure with which the missing values are imputed.  -->

@@

<!-- based on a model of the observed databy means of an iterative algorithm  -->
<!-- is the use of MICE: multiple imputation by chained equations. MICE is a type of multiple imputation procedure--a framework proposed by @rubin87--in which missing datapoints are 'imputed' (i.e., filled in) by means of an iterative algorithm (chained equations). Imputed values are obtained by modeling the (distribution of the) missing data conditional on the observed data. The technique is therefore also called 'fully conditional specification' (FCS). -->
<!-- several times, and inferences are aggregated across imputed datasets.  -->
<!-- and statistical inference is performed on each imputed dataset. Variation between the results refelect uncertainty due to missingness  -->
<!-- CE: The imputation model can be a joint model for all variables, or several univariate conditional distributions (fully conditional specification). FCS is more flexible and "avoids the problem of specifying an appropriate joint imputation distribution and replaces this by the selection of appropriate univariate conditional distributions".  -->

@@

**(There is no scientific consensus on the convergence properties of MI algorithms [@taka17]. Some default MICE techniques (e.g., 'predictive mean modeling') might not yield converged states at all [@murr18]. Therefore, algorithmic convergence should be monitored carefully.)**

The recommended practice to monitor convergence is to visually inspect imputation chains for signs of non-convergence. This practice is insuffcient on two counts: 1) it may be challenging to the untrained eye [@buur18, $\S$ 6.5.2], and 2) only severely pathological cases of non-convergence may be diagnosed. Therefore, a quantitative, diagnostic evaluation of convergence would be preferred.

Iterative imputation algorithms such as MICE are MCMC methods, and convergence of MCMC algorithms is not from a scalar to a point but from one distribution to another. Therefore, generated values will vary even after convergence. The aim of convergence diagnostics for MCMC methods is thus not to establish the point at which convergence is reached, but to diagnose signs of  non-convergence.
<!-- Therefore we can only diagnose non-convergence, not convergence [@hoff09]. "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->

Several MCMC convergence diagnostics exist, but it is not known whether these are appropriate for MICE. The application of convergence diagnostics to MI methods has not been systematically studied [@buur18]. The open questions are, e.g.: How can non-convergence be diagnosed? How many iterations are sufficient/needed? What are the effects of non-convergence on estimates, predictions and inference? What is an appropriate non-convergence diagnostic? Are the default number of iterations sufficient? The defaults of common MI software: SPSS = 10 [(link)](https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html), Mplus = 100?? [(link)](https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748), Stata = 10 [(link)](https://www.stata.com/manuals13/mi.pdf, p. 139), Amelia = NA, because resampling, not convergence [(link)](https://cran.r-project.org/web/packages/Amelia/Amelia.pdf), en MI = 30 [(link)](https://cran.r-project.org/web/packages/mi/mi.pdf).

@@

Let $y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample. The data value of unit $i$ ($i = 1, \dots, n$) on variable $j$ ($j = 1, \dots, p$) may be either observed or missing. The collection of observed data values in $y$ is denoted by $y_{obs}$; the missing part of $y$ is referred to as $y_{mis}$. (**This is not only notation, but theory too:** For each datapoint in $y_{mis}$, we sample $M \times T$ times plausible values, where $M$ is the number of imputations ($m = 1, \dots, M$) and $T$ is the number of iterations ($t = 1, \dots, T$).) The collection of samples between the initial value (at $t=1$) and the final imputed value (at $t=T$) will be referred to as an 'imputation chain'.

**Add: missingness percentage = proportion of cases with one or more missing values times 100 **
<!-- The number of imputations is denoted by $m$ (i.e., the number of times each datapoint in $y_{mis}$ is filled in). The aim of MI is to impute each datapoint in $y_{mis}$ $m$ times. Because of the iterative nature of the MICE algorithm -->

@@

**(Start with current guidelines, then continue with possible quant measures of the same things from MCMC lit.)** In iterative algorithmic procedures (Markov chain Monte Carlo methods; e.g., `mice` algorithms or Gibbs samplers) non-convergence may be present as non-stationarity within chains (i.e., trending), or as slow mixing between chains (i.e., no intermingling (**source??**)). (**add example/guidelines visual inspection here?**)

The stationarity component of convergence may be evaluated with autocorrelation [$AC$; @scha97; @gelm13], numeric standard error [or `MC error'; @gewe92], and Raftery and Lewis's [-@raft91] procedure to determine  the effect of trending within chains (**or the chain length required to cancel out trending?**).

The mixing component can be assessed with the potential scale reduction factor $\widehat{R}$ [a.k.a. `Gelman-Rubin statistic'; @gelm92]. With an adapted version of $\widehat{R}$, proposed by @veht19, the stationarity component of convergence might be evaluated as well. This would make $\widehat{R}$ a general convergence diagnostic. The application of $\widehat{R}$ to assess stationarity has not been thoroughly investigated. Therefore, this study employs both $\widehat{R}$ and autocorrelation to investigate convergence, as recommended by [@cowl96, p. 898].

@@

Note that all of these methods evaluate the convergence of univariate scalar summaries (e.g., chain means or variances). These convergence diagnostics cannot diagnose convergence of multivariable statistics (i.e., relations between scalar summaries). @buur18 proposed to implement multivariable evaluation through eigenvalue decomposition [@mack03]. This method is outside of the scope of the current study (**not anymore?**).

@@

We can interpret $AC$-values as a measure of stationarity. If $AC$-values are close to zero, there is no dependence between subsequent samples within imputation chains. Positive $AC$-values, however, indicate recurrence. If $\theta$-values of subsequent iterations are similar, trending may occur. Negative $AC$-values show no threat to the stationarity component of convergence. On the contrary even---negative $AC$-values indicate that $\theta$-values of subsequent iterations diverge from one-another, which may increase the variance of $\theta$ and speed up convergence. As convergence diagnostic, the interest is therefore in positive $AC$-values. Moreover, the magnitude of $AC$-values may be evaluated statistically, but that is outside of this note's scope.

Negative $AC$-values indicate divergence within imputation chains. Subsequent sampled values within each imputation chain are less alike. It may not be wise (**why? and how would we know? e.g., wider distrubutions? didn't seem like it...**) to terminating the algorithm when AC is negative. **High $AC$-values are implausible in MI procedures. That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains.**

@@

Based on an empirical finding [@lace07], we hypothesize that $\widehat{R}$ will over-estimate non-convergence of MI algorithms. The threshold of $\widehat{R} < 1.01$ will then be too stringent for diagnosing convergence. This over-estimation may, however, be diminished because $\widehat{R}$ can falsely diagnose convergence if initial values of the algorithm are not appropriately over-dispersed [@broo98, p. 437]. In `mice`, initial values are chosen randomly from the observed data. Therefore, we cannot be certain that the initial values are over-dispersed. We expect this to have little effect on the hypothesized performance of $\widehat{R}$. **(Add actual hypothesis.)** No hypothesis was formulated about the performance of $AC$ as convergence diagnostic.

@@

```{r eval=FALSE}
# pseudo-code of simulation 
simulate data 
for (number of simulation runs from 1 to 1000)
  for (number of iterations from 1 to 100)
    create missingness
    impute the missingness
    compute convergence diagnostics
    perform analysis
    pool results
    compute simulation diagnostics
aggregate convergence and simulation diagnostics
```

@@

**Add the function `mvtnorm::rmvnorm()` back in** use citation("mvtnorm")

The complete data is 'amputed' once for each simulation repetition with function `mice::ampute()`. (**change this:** The missingness is univariate, and the probability to be missing is the same for all four variables, namely 20\% (`prop = 0.8, mech = "MCAR"`). This leaves 20\% of the rows completely observed). **(Add different percentages missingness??)** This study only considers only an MCAR missingness mechanism. Therefore, results may not be extrapolated to other missing data problems. Proper performance of the convergence diagnostics under MCAR is necessary but not sufficient to demonstrate appropriateness of $\widehat{R}$ and $AC$ as convergence diagnostics. This is just a proof of concept.

@@

## Univariate estimates and convergence diagnostics (theta = mean and theta = variance)

@@

There is a clear bias in the regression estimates in simulation conditions where the number of iterations is smaller than four. In simulation conditions where $T > 5$ there is little to no bias in the estimated regression coefficients. In most simulation conditions nominal coverage is obtained (i.e., coverage rates of .95) for the confidence intervals around the regression coefficients. Conditions with only one or two iterations show some undercoverage. Since confidence interval width is stable across conditions, the undercoverage may be attributed to the bias in the estimated regression coefficients.

@@

@@

```{r, include = FALSE}
load("../Results/complete.Rdata")
source("../2.CreateFigures.R")
```

<!-- ```{r} -->
<!-- bias_means_plot + R_means_plot + AC_means_plot + plot_layout(nrow = 3) + plot_annotation(title = "Convergence of means, 75% missingness") -->
<!-- ``` -->

```{r, echo = FALSE}
#load("../Results/combined.Rdata")

dat %>% ggplot(aes(x = T, y = bias.R.s, color = as.factor(miss))) + 
  geom_point() +
  geom_line() +
  ylab(expression(paste("Bias in ", hat(R ^ 2)))) +
  labs(colour = "Legend: missingness percentage") +
  plot_annotation(title = "The effect of missingness percentage, an example")

@@

# Results [OLD, JUST FOR REFERENCE!]

## Convergence Diagnostics

**(Pair convergnce diagnostics with respective estimates, not split between convergence and simulation diagnostics)**

It is apparent that there is a relation between the number of iterations per simulation condition ($T$) and the convergence diagnostics. Generally speaking, conditions with longer imputation chains (higher $T$) coincide with less signs of non-convergence ($\widehat{R}$-values approach one, and $AC$-values approach zero). 
<!-- We see more or less equivalent trends for chain means versus chain variances. We, therefore, only discuss the $\widehat{R}$ and $AC$-values of chain means.  -->

Figure 2A shows that $\widehat{R}$-values generally decrease with increasing imputation chain lengths. The decline stabilizes somewhere between the simulation conditions $T=30$ and $T=50$. The downward trend is most pronounced for $T=3$, and between $T = 5$ and $T = 10$. In the intervening conditions ($3 \leq T \leq 5$), however, we observe a steep increase in $\widehat{R}$-values. This increase implies that non-convergence is under-estimated, and convergence should not be diagnosed at this point (**is that true???**). Based on the conventional threshold $\widehat{R} < 1.2$, we would falsely diagnose convergence at $T=3$. According to the widely used threshold $\widehat{R} < 1.1$, convergence would be diagnosed for conditions where $T>9$. If we use the recently recommended threshold $\widehat{R} < 1.01$, we would conclude that convergence is reached in none of the simulation conditions.

The $AC$-values displayed in Figure 2B are almost uniformly increasing as a function of $T$. The only $AC$-value that deviates from this observation is for $T=2$. The lowest $AC$-value is obtained for $T=3$. %At $T=5$, the $AC$-value reaches the level observed at $T=2$. 
The gradual increase plateaus between $T=10$ and $T=30$. $AC$-values in conditions where $T>70$ are indifferentiable from zero, indicating stationarity. We see an initial decrease between $T=2$ and $T=3$. Simulation conditions where $T>5$ have $AC$-values greater than $T=2$ (**weird wording**).

According to this diagnostic, none of the simulation conditions show signs of non-convergence, since we only observe negative or zero $AC$-values.

Taken together, we see that $T>3$ is the minimal requirement to diagnose convergence ($\widehat{R} < 1.2; AC \leq 0$). This threshold is, however, not sufficient, since we overlook the increase in $\widehat{R}$-values up-to conditions where $T>5$, and the convergence diagnostics only reach stability at $T>20$.

## Simulation Diagnostics

We use average bias, average confidence interval width, and coverage rate as performance measures to evaluate $\widehat{R}$ and $AC$ (**not the goal! we want to show the effects of non-conv, not just see how to diagnose it.**). We make the general observation that the simulation diagnostics behave as theorized for most simulation conditions (bias around zero, stable confidence interval widths, nominal coverage rate at 95\%).

Figure 3A shows that bias is fairly stable across simulation conditions. The condition $T=1$ clearly deviates from this trend with a negative bias. The bias for $T=2$ is below average, but within the range of fluctuations. Conditions where $T>3$ can be diagnosed as unbiased, but the average bias across iterations (as shown by a flat Loess line) only reaches stability at $T=20$ (**Loess might not be relevant, check and remove accordingly!**).

CIW is only clearly divergent in the simulation condition $T=1$, see Figure 3B. Only conditions where $T>3$ are therefore considered sufficient. (**Possibly irrelevant Loess:** These conditions have similar CIWs, but across iterations stability is not established (i.e., the Loess line is never completely flat within the 100 simulation conditions).)

The empirical coverage rate across repetitions seems more or less stable for conditions where $T>1$, see Figure 3C. We see some over-coverage at $T=2$, but that is better than under-estimating the variance of $\bar{Q}$. On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% [@neym34] (**not anymore??**). (**Loess:** Similar to the CIWs, there is some trending across iterations (i.e., the Loess line is never flat)).

From the simulation diagnostics, we observe that unbiased estimates with nominal coverage rates were obtained in conditions where $T>3$. This suggests that as little as four iterations may be sufficient for the MI algorithm under the current circumstances.

In short, there is a discrepancy between what the convergence diagnostics and what the performance measures indicate. While $T=4$ seems sufficient with respect to simulation quantities, the condition where $T=4$ resulted in the `one-but-worst' values of both convergence diagnostics. Complete algorithmic convergence as indicated by $\widehat{R}$ and autocorrelation is not reached in conditions where $T<20$.

>>>>

---
title: Missing the Point$\colon$ Non-Convergence in Iterative Imputation Procedures
runninghead: Oberman
author:
- name: H. I. Oberman
  num: 1
address:
- num: 1
  org: Department of Methodology and Statistics, Utrecht University, Utrecht, The Netherlands
corrauth: "Hanne Oberman, Sjoerd Groenman building, Utrecht Science Park, Utrecht, The Netherlands."
email: h.i.oberman@uu.nl
abstract: "(**Rewrite:**) Multiple imputation by chained equations (mice) is a widely used tool to accommodate missing data. While empirical evidence supports the validity of inferences obtained using mice, there is no consensus on the convergence properties of the method. This paper provides insight into non-convergence of mice algorithms."
keywords: MICE; convergence 
classoption:
  - Royal
  - times
bibliography: thesis.bib
bibliographystyle: sageh
output:
  rticles::sage_article:
    keep_tex: yes
---

@@

**Note for peer review: things written in bold are reminders for myself or things that I am unsure off still. Also, I just decided to change things majorly in my results section. So right now, the figures do not correspond to the text. Please only skim that section, because I am definitely going to rewrite it. Thanks already! - Hanne**

@@

<!-- (**or** Everybody who works with person data is likely to run into a missing data problem.) -->
Missing data problems are ubiquitous and pervasive in the social and behavioral sciences. If a dataset contains just one missing value for a variable of interest, statistical inferences are undefined and will not produce any result. Incomplete observations are therefore ignored by default in many statistical packages (i.e., list-wise deletion is employed). Unfortunately, this *ad hoc* solution may yield wildly invalid results [@buur18]. An alternative is to 'impute' (i.e., fill in) every missing value in an incomplete dataset. With imputation techniques, one or several completed datasets are created, on which statistical inferences can be performed. The case with several completed datasets is known as 'multiple imputation' [MI; @rubin76], and requires an additional step to pool the results across imputations. By applying Rubin's rules, the pooled result will reflect the uncertainty in the data due to missingness. Under many circumstances, this yields an unbiased and confidence valid estimate of the inference on the true--but missing--data [@buur18]. A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'[@mice]. MICE is an iterative algorithmic procedure to draw imputations from the posterior predictive distribution of the missing values. This introduces a potential threat to the validity of the imputations: what if the algorithm has not converged? Are the implications then to be trusted? And can we rely on the inference obtained on the completed data? These are all open questions, because the convergence properties of iterative imputation algorithms have not been systematically studied [@buur18]. Moreover, there is no scientific consensus on the convergence properties of MI algorithms [@taka17]. Some default MICE techniques (e.g., 'predictive mean modeling') might not yield converged states at all [@murr18]. Therefore, algorithmic convergence should be monitored carefully.

<!-- In this paper, we investigate the convergence properties of iterative imputation procedures by means of model based simulation in `R` [@R].  -->

@@

<!-- by which  That is, unless ad hoc solutions are employed (e.g., list-wise deletion), which can yield invalid results.  contains missing values, statistical analyses cannot be performed without  only be performed a dataset contains missing values, statistical analyses can statistical inferences cannot be performed   -->
<!-- At some point, any social scientist  -->
<!-- conducting statistical analyses  --> <!-- will run into a missing data problem [@alli02].  -->
<!-- Missingness is problematic because statistical inference cannot be performed on the whole data without employing \emph{ad hoc} solutions (e.g., list-wise deletion), which may yield wildly invalid results [@buur18]. A popular method to obtain statistically valid results despite of missing data is 'multiple imputation' [MI; @rubin76]. The multiple imputation method can be roughly divided into 3 steps: First, each missing value in an incomplete dataset is 'imputed' (i.e., filled in) several times by some algorithmic procedure. Next, the analysis of scientific interest is performed on each of the completed datasets. And finally, the results of these analyses are pooled into a single aggregated statistic.  -->

@@

<!-- The theoretical foundation of these steeps was laid by Rubin [-@rubin76, -@rubin87]. For an accessible and comprehensive introduction to MI from an applied perspective, see e.g. @buur18. In this paper, we focus on the first of these steps: the algorithmic procedure with which the missing values are imputed.  -->

@@

<!-- based on a model of the observed data by means of an iterative algorithm  -->
<!-- is the use of MICE: multiple imputation by chained equations. MICE is a type of multiple imputation procedure--a framework proposed by @rubin87--in which missing datapoints are 'imputed' (i.e., filled in) by means of an iterative algorithm (chained equations). Imputed values are obtained by modeling the (distribution of the) missing data conditional on the observed data. The technique is therefore also called 'fully conditional specification' (FCS). -->
<!-- several times, and inferences are aggregated across imputed datasets.  -->
<!-- and statistical inference is performed on each imputed dataset. Variation between the results reflect uncertainty due to missingness  -->
<!-- CE: The imputation model can be a joint model for all variables, or several univariate conditional distributions (fully conditional specification). FCS is more flexible and "avoids the problem of specifying an appropriate joint imputation distribution and replaces this by the selection of appropriate univariate conditional distributions".  -->

@@

The recommended practice to monitor convergence is to visually inspect imputations for signs of non-convergence. This practice is insufficient on two counts: 1) it may be challenging to the untrained eye, and 2) only severely pathological cases of non-convergence may be diagnosed [@buur18, $\S$ 6.5.2]. Therefore, a quantitative, diagnostic evaluation of convergence would be preferred.

Iterative imputation algorithms such as MICE are Markov chain Monte Carlo (MCMC) methods, which means that convergence is not from a scalar to a point but from one distribution to another. The values generated by the algorithm (e.g., imputed values) will vary even after convergence. Therefore, the aim of convergence diagnostics for MCMC methods should not be to establish the point at which convergence is reached, but to monitor signs of  non-convergence. Several convergence diagnostics for MCMC methods exist, but it is not known whether these are appropriate for MICE.

<!-- Therefore we can only diagnose non-convergence, not convergence [@hoff09]. "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->

Main question: How can non-convergence be diagnosed? Sub-questions: Are common MCMC non-convergence diagnostics appropriate for MICE? And if so, which threshold should be used? How many iterations are sufficient/needed to reach convergence? Are the default number of iterations sufficient (i.e., 5 in mice, 10 in SPSS and Stata, 30 in mi)? What are the effects of continued iterations on estimates, predictions and inferences? And finally, do the answers differ with varying missingness proportions?

<!-- The defaults of common MI software: SPSS = 10 [(link)](https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html), Mplus = 100?? [(link)](https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748), Stata = 10 [(link)](https://www.stata.com/manuals13/mi.pdf, p. 139), Amelia = NA, because re-sampling, not convergence [(link)](https://cran.r-project.org/web/packages/Amelia/Amelia.pdf), en MI = 30 [(link)](https://cran.r-project.org/web/packages/mi/mi.pdf). -->

@@

Let $y$ denote an $n \times p$ matrix containing the data values on $p$ variables for all $n$ units in a sample. The data value of unit $i$ ($i = 1, \dots, n$) on variable $j$ ($j = 1, \dots, p$) may be either observed or missing. The collection of observed data values in $y$ is denoted by $y_{obs}$; the missing part of $y$ is referred to as $y_{mis}$. For each datapoint in $y_{mis}$, we sample $M \times T$ times plausible values, where $M$ is the number of imputations ($m = 1, \dots, M$) and $T$ is the number of iterations ($t = 1, \dots, T$). The collection of samples between the initial value (at $t=1$) and the final imputed value (at $t=T$) will be referred to as an 'imputation chain'. **Add: missingness percentage = proportion of cases with one or more missing values **
<!-- The number of imputations is denoted by $m$ (i.e., the number of times each datapoint in $y_{mis}$ is filled in). The aim of MI is to impute each datapoint in $y_{mis}$ $m$ times. Because of the iterative nature of the MICE algorithm -->

@@

The two challenges for convergence of iterative algorithms are mixing and stationarity [@gelm13], see Figure 1. Without mixing, imputation chains do not intermingle nicely. Without stationarity, there is trending within imputation chains. While the current practice is to inspect traceplots for signs of non-mixing and non-stationarity, diagnostic evaluation of these two challenges would be preferred. Non-stationarity within chains may be diagnosed with e.g., autocorrelation [$AC$; @scha97; @gelm13], numeric standard error ['MC error'; @gewe92], or Raftery and Lewis's [-@raft91] procedure to determine the effect of trending on the precision of estimates. A widely used diagnostic to monitor mixing between chains is the potential scale reduction factor $\widehat{R}$ ['Gelman-Rubin statistic'; @gelm92]. With a recently proposed adaptation, $\widehat{R}$ might also serve to diagnose non-stationarity, but this has not yet been thoroughly investigated [@veht19]. Therefore, mixing and stationarity will be evaluated separately in this study. As recommended [e.g., @cowl96, p. 898], $AC$ and $\widehat{R}$ will be used.

<!-- The current practice in diagnosing non-convergence is to inspect traceplots for signs of non-mixing and non-stationarity.   -->

<!-- The mixing component can be assessed with the potential scale reduction factor $\widehat{R}$ [a.k.a. `Gelman-Rubin statistic'; @gelm92]. With an adapted version of $\widehat{R}$, proposed by @veht19, the stationarity component of convergence might be evaluated as well. This would make $\widehat{R}$ a general convergence diagnostic. The application of $\widehat{R}$ to assess stationarity has not been thoroughly investigated. Therefore, this study employs both $\widehat{R}$ and autocorrelation to investigate convergence, as recommended by [@cowl96, p. 898]. -->

@@

<!-- Note that all of these methods evaluate the convergence of univariate scalar summaries (e.g., chain means or variances). These convergence diagnostics cannot diagnose convergence of multivariable statistics (i.e., relations between scalar summaries). @buur18 proposed to implement multivariable evaluation through eigenvalue decomposition [@mack03]. This method is outside of the scope of the current study (**not anymore?**). -->

@@

We can interpret $AC$-values as a measure of stationarity. If $AC$-values are close to zero, there is no dependence between subsequent samples within imputation chains.

Positive $AC$-values indicate recurrence. If $\theta$-values of subsequent iterations are similar, trending may occur. Negative $AC$-values show no threat to the stationarity component of convergence. On the contrary even---negative $AC$-values indicate that $\theta$-values of subsequent iterations diverge from one-another, which may increase the variance of $\theta$ and speed up convergence. As convergence diagnostic, the interest is therefore in positive $AC$-values. Moreover, the magnitude of $AC$-values may be evaluated statistically, but that is outside of this note's scope.

Negative $AC$-values indicate divergence within imputation chains. **Subsequent sampled values within each imputation chain are less alike. High $AC$-values are implausible in MI procedures. That is, the randomness induced by the MI algorithm effectively mitigates the risk of dependency within chains.**

@@

Based on an empirical finding [@lace07], we hypothesize that $\widehat{R}$ will over-estimate non-convergence of MI algorithms. The threshold of $\widehat{R} < 1.01$ will then be too stringent for diagnosing convergence. This over-estimation may, however, be diminished because $\widehat{R}$ can falsely diagnose convergence if initial values of the algorithm are not appropriately over-dispersed [@broo98, p. 437]. In `mice`, initial values are chosen randomly from the observed data. Therefore, we cannot be certain that the initial values are over-dispersed. We expect this to have little effect on the hypothesized performance of $\widehat{R}$. **(Add actual hypothesis: we thought that mice would converge sooner than $\widehat{R}$ would indicate that it did)** No hypothesis was formulated about the performance of $AC$ as convergence diagnostic.

@@

```{r, eval=FALSE, echo=TRUE}
# pseudo-code of simulation 
simulate data 
for (number of simulation runs from 1 to 1000)
  for (number of iterations from 1 to 100)
    create missingness
    impute the missingness
    compute convergence diagnostics
    perform analysis
    pool results
    compute simulation diagnostics
aggregate convergence and simulation diagnostics
```

**(Add step for different missingness proportions.)**

@@

**(Add the function `mvtnorm::rmvnorm()` back in somewhere here? If yes, use citation("mvtnorm").)**

The complete data is 'amputed' once for each simulation repetition with function `mice::ampute()`. (**change this to reflect current simulation set-up with 5, 25, 50, 75, and 95% of cases having missing data:** The missingness is univariate, and the probability to be missing is the same for all four variables, namely 20\% (`prop = 0.8, mech = "MCAR"`). This leaves 20\% of the rows completely observed). This study only considers only an MCAR missingness mechanism. Therefore, results may not be extrapolated to other missing data problems. Proper performance of the convergence diagnostics under MCAR is necessary but not sufficient to demonstrate appropriateness of $\widehat{R}$ and $AC$ as convergence diagnostics. This is just a proof of concept.

@@

```{r, include = FALSE}
load("../Results/test.Rdata")
source("../2.CreateFigures.R")
```

```{r}
mean_bias + plot_annotation(title = "Bias in mean estimate, convergence of chain means")
mean_Rh 
mean_ACF + plot_annotation(title = "Dotted lines are old AC computation, solid are new.")
```

## Univariate estimates and convergence diagnostics

@@

There is a clear bias in the regression estimates in simulation conditions where the number of iterations is smaller than four. In simulation conditions where $T > 5$ there is little to no bias in the estimated regression coefficients. In most simulation conditions nominal coverage is obtained (i.e., coverage rates of .95) for the confidence intervals around the regression coefficients. Conditions with only one or two iterations show some under-coverage. Since confidence interval width is stable across conditions, the under-coverage may be attributed to the bias in the estimated regression coefficients.

@@

```{r}
est_bias + plot_annotation(title = "Bias in regression estimate, coverage rate, \n bias in variance explained")
est_cov
Rsq_bias 
```

@@

@@

<!-- # Results [OLD, JUST FOR REFERENCE!] -->

<!-- ## Convergence Diagnostics -->

<!-- **(Pair convergence diagnostics with respective estimates, not split between convergence and simulation diagnostics)** -->

<!-- It is apparent that there is a relation between the number of iterations per simulation condition ($T$) and the convergence diagnostics. Generally speaking, conditions with longer imputation chains (higher $T$) coincide with less signs of non-convergence ($\widehat{R}$-values approach one, and $AC$-values approach zero).  -->
<!-- <!-- We see more or less equivalent trends for chain means versus chain variances. We, therefore, only discuss the $\widehat{R}$ and $AC$-values of chain means.  --> -->

<!-- Figure 2A shows that $\widehat{R}$-values generally decrease with increasing imputation chain lengths. The decline stabilizes somewhere between the simulation conditions $T=30$ and $T=50$. The downward trend is most pronounced for $T=3$, and between $T = 5$ and $T = 10$. In the intervening conditions ($3 \leq T \leq 5$), however, we observe a steep increase in $\widehat{R}$-values. This increase implies that non-convergence is under-estimated, and convergence should not be diagnosed at this point (**is that true???**). Based on the conventional threshold $\widehat{R} < 1.2$, we would falsely diagnose convergence at $T=3$. According to the widely used threshold $\widehat{R} < 1.1$, convergence would be diagnosed for conditions where $T>9$. If we use the recently recommended threshold $\widehat{R} < 1.01$, we would conclude that convergence is reached in none of the simulation conditions.  -->

<!-- The $AC$-values displayed in Figure 2B are almost uniformly increasing as a function of $T$. The only $AC$-value that deviates from this observation is for $T=2$. The lowest $AC$-value is obtained for $T=3$. %At $T=5$, the $AC$-value reaches the level observed at $T=2$.  -->
<!-- The gradual increase plateaus between $T=10$ and $T=30$. $AC$-values in conditions where $T>70$ are indifferentiable from zero, indicating stationarity. We see an initial decrease between $T=2$ and $T=3$. Simulation conditions where $T>5$ have $AC$-values greater than $T=2$ (**weird wording**).  -->

<!-- According to this diagnostic, none of the simulation conditions show signs of non-convergence, since we only observe negative or zero $AC$-values.  -->

<!-- Taken together, we see that $T>3$ is the minimal requirement to diagnose convergence ($\widehat{R} < 1.2; AC \leq 0$). This threshold is, however, not sufficient, since we overlook the increase in $\widehat{R}$-values up-to conditions where $T>5$, and the convergence diagnostics only reach stability at $T>20$.  -->

<!-- ## Simulation Diagnostics -->

<!-- We use average bias, average confidence interval width, and coverage rate as performance measures to evaluate $\widehat{R}$ and $AC$ (**not the goal! we want to show the effects of non-conv, not just see how to diagnose it.**). We make the general observation that the simulation diagnostics behave as theorized for most simulation conditions (bias around zero, stable confidence interval widths, nominal coverage rate at 95\%).  -->

<!-- Figure 3A shows that bias is fairly stable across simulation conditions. The condition $T=1$ clearly deviates from this trend with a negative bias. The bias for $T=2$ is below average, but within the range of fluctuations. Conditions where $T>3$ can be diagnosed as unbiased, but the average bias across iterations (as shown by a flat Loess line) only reaches stability at $T=20$ (**Loess might not be relevant, check and remove accordingly!**).  -->

<!-- CIW is only clearly divergent in the simulation condition $T=1$, see Figure 3B. Only conditions where $T>3$ are therefore considered sufficient. (**Possibly irrelevant Loess:** These conditions have similar CIWs, but across iterations stability is not established (i.e., the Loess line is never completely flat within the 100 simulation conditions).)   -->

<!-- The empirical coverage rate across repetitions seems more or less stable for conditions where $T>1$, see Figure 3C. We see some over-coverage at $T=2$, but that is better than under-estimating the variance of $\bar{Q}$. On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% [@neym34] (**not anymore??**). (**Loess:** Similar to the CIWs, there is some trending across iterations (i.e., the Loess line is never flat)). -->

<!-- From the simulation diagnostics, we observe that unbiased estimates with nominal coverage rates were obtained in conditions where $T>3$. This suggests that as little as four iterations may be sufficient for the MI algorithm under the current circumstances. -->

<!-- In short, there is a discrepancy between what the convergence diagnostics and what the performance measures indicate. While $T=4$ seems sufficient with respect to simulation quantities, the condition where $T=4$ resulted in the `one-but-worst' values of both convergence diagnostics. Complete algorithmic convergence as indicated by $\widehat{R}$ and autocorrelation is not reached in conditions where $T<20$.   -->
